Noisy Work: A Review of The Uncertainty Visualisation Literature

### 2024-02-27 paper.qmd
word vomit of great ideas and updated edit notes

---
title: "Noisy Work: A Review of The Uncertainty Visualisation Literature"
author: Harriet Mason
bibliography: references.bib
date: last-modified
toc: true
number-sections: true
format: pdf
editor_options: 
  chunk_output_type: console
---
# Background
## Introduction

**Aside: I have not touched this section since the confirmation so it would be best to ignore it**

From entertainment choices to news articles to insurance plans, the modern citizen is so over run with information in every aspect of their life it can be overwhelming. So overwhelming in fact, that it is not uncommon for people to become so overwhelmed by the information they refuse to engage with it at all. In this overflow of information, tools that can effectively summarise information down into simple and clear ideas become more valuable. While websites that summarise insurance plans or hotel options down into a clear table make a living off it, few tools are as powerful or effective as a visualisation. Information visualisations remain one of the most powerful tools for fast and reliable science communication. Effectively leveraging visualisations to make them more effective is therefore key in facilitating simple and clear communication of science and other ideas.

@@

- @Suh2023 found that the main issue with deployment of AI was communication between data science and subject matter experts, since subject matter experts didn't want technical presentations that didn't explain exactly when and how a model would fail.

@@

- industrial accidents have a "tollerable" risk chart based on risk and number of fatalities @Spiegelhalter2017

## Similar work
There are a wealth of uncertainty communication and visualisation literature reviews that already exist. Interestingly they all seem to have identified similar pitfalls or fallen into one themselves.

@@

## Coloquial Definitions
**Aside: I am not sure how to collect these together and show that they are highly varied. This might be a situation for the tidytext clustering or something. There is an additional problem is that the definition for uncertainty is completely unrelated to the uncertainty that is quantified in the graphics. I think the tidytex can help with the definitions being all over the place, but I am not sure how to identify the lack of correlation between the colloquial definitions of uncertainty and the mathematical quantificaitons.**

The most common definition of uncertainty used in papers that discuss visualising it are colloquial. Most definitions make some nod towards variance, precision, trust, error, missing values or another related concepts. Since *uncertainty*, unlike *probability*, does not have a strict definition that is established in mathematics, there is large variety in its usage. Some consider it to be synonymous with probabiltiy, some consider it to be an encompassing umbrella term of which probability is only an element, and some consider it to by something else entirely. This lack of a definition leads to uncertainty being swapped out freely with other vague words such as ambiguity and risk. Definitions presented in papers that discuss uncertainty visualisation want it to be encompassing and include everything a layperson might consider when trying to consider "uncertainty". The issue with these broad definitions is is that in order to visualize something, it needs to be quantified, and in order to quantify something, we need to know exactly what it is we are counting. This discrepancy appears regularly in the uncertainty visualisation literature. Often a vague definition such as "uncertainty is anything that isn't deterministic", or even a refusal to define uncertainty at all in an experiments introduction will be accompanied by a highly specific quantification of uncertainty as a probability density distribution of a forecast when providing the visualisation. Ironically it seems that many of the papers that refer to themselves as "uncertainty" papers are just suffering from a form of uncertainty, a lack of precision.[^1]

@@

## Taxonomy Definitions
### Taxonomies of Uncertainty
**Aside: I actually think I agree with the sentiment in the grammar of graphics that a taxonomy is more of an opinion and we can reclassify it and is therefore not thorough enough to remain unchanged and be build upon. That being said, if this is a literature review and all existing literature use written definitions or taxnomies, that is what I have to work with. The larger taxonomy can be used to show that the other taxonomies fit within it, but it is in of itself not tied to anything except work that paper itself identifies as transient.**

@@

**Aside: This figure is not going to be in the final version of the paper. It is here as a placeholder. Ideally, I would replace with with a graphic that shows where all these taxonomies land in the location, level, nature taxonomy (since it is encompassing) and show they are very inconsistent and a new definition is pulled out for almost every paper (even in statistics papers)**

@@

*I need to fix the taxonomy illustration because inputs and model uncertainty need to be swapped*

![Depicts an illustration of the taxonomy described in @utypo. From right to left the drawing shows the location, level and nature of uncertainty with examples of that category underneath. A specific source of uncertainty from the location can be mapped to a level of ignorance that can increase or decrease (i.e. moving up or down the green line) depending on the nature of the uncertainty. Identifying the location, level and nature of your uncertainty allows you to better understand it. ](taxonomyvis.jpeg){#fig-taxonomy}

**Aside: I had this here to explain why these are the considerations people make in statistics and how they relate to elements of the taxonomy to try and provide some justification, but If I am just going to show the definitions are inconsitent, I might get rid of this section. I also largely included it so you guys can see what the taxonomies are considering**

@@

Mapping uncertainty into the 3D space defined by location, level, and nature can expand beyond the narrow case of statistical modelling it was defined for. *@___* included "consensus" as a taxonomy element, however it was not included in the final definition, but it can easily be added as a location that would align with the communication stage of the data pipeline. Several authors [*@_+_@Spiegelhalter2017*] considered the method by which uncertainty was quantified (e.g. by a single value, a function, simulated outcomes, etc), which can be be thought of as a more precise categorisation of the level axis. This taxonomy also ignores contextual information about the data, such as spatial or temporal considerations. This flexibility means that the use of this framework is not limited by a specific case and should be used going forward to define what aspect of the uncertainty space a contribution exists in.

@@

### Taxonomies of Visual Uncertainty
There are some taxonomies of visual uncertainty. These taxonomies seem to seek to define "uncertainty visualisation" and a combined term rather than as two separate words. We largely disagree with this conceptualization of uncertainty visualisations as it blurs the line between the mathematical estimations and assumptions (the uncertainty part) and the visual depiction of those mathematical objects (the visualisation part). While we will not focus on these taxonomies, it is still important to mention them and highlight the reasons they are not good practice.

@@

**Aside: I should add something in about the actual quantified uncertainty vs the discussed uncertanty thing (i.e. what is defined with words and what is defined with maths)**

## Contextual Information
Some uncertainty work, specifically uncertainty visualisation, focus on the context surrounding the uncertainty. These are things like spatial data, temporal data, category data, etc. While other discussions of uncertainty may consider this distinction more relevant to their work, here we will consider this contextual information. Contextual information is important for interpretation and understanding of graphics, but it does not in of itself generate uncertainty. This is not unique to the properties of spatial information, other statistics information that is important for modelling and assumptions, such as non-linearity, class levels, etc do not contribute to uncertainty. Operations on data, inherrent randomness, and assumptions that create bias generate uncertainty, which is what we are talking about here. A confidence interval is a confidence interval whether it came from spatial data, temporal, or cross sectional data.

@@

# General Attitude Towards Uncertainty (Uncertainty Communication Boradly)
It would be impossible to discuss uncertainty visualisation without mentioning the broader conversation around uncertainty communication. The conversation around uncertainty communication is often centered around the psychology is communicating risks.

@@

### Extra notes for this section
- People believe in the law of small numbers but dont believe in the lar of large numbers @Hullman2016
- experiencing an event makes your percieved probability higher than if you see a description of it @Hullman2016
- Frequency thing means that more effective communication to participants in a study means more effective science and we have a roll on effect @Hullman2016
- Probability judgements are less relative than other psychophysical judgements, 0%, 50% and 100% all have well understood meanings. Subjective probability distributions are spontaneous, and subjects are constructing a distribution upon being asked rather than articulating a fully formed mental representation of their beliefs. Therefore it is subject to heuristics. consider anchoring a scale with farmiliar probabilities . not sure if being asked for the entire distribution helps, e.g. the graphical one reduces noise but asking people for an interval estimation leads to overprecision and overconfidence @Hullman2016
- believed that offering enough decisions with varying rewards you can observe the true subjective probability, however whether or not probability is the right representation of subjective uncertainty is a topic of debate @Hullman2016
- she also discusses graphical interface (users have to draw the probability distribution) and framing probabilities as natural frequencies are better choices because they reduce noise and stuff.@Hullman2016
- Consider incorperating utility functions. Include "probability-coherence" checks @Hullman2016
- evidence that people do not follow the axioms of probability when conveying subjective probabilities. How do you deal with this as an analyst (e.g. if the probabilities do not sum to 1 @Hullman2016

@Spiegelhalter2017
we consider a summary table as a visualisation
illuminate graphics with words and numbers, 
design graphics to allow a part-to-whole comparison on an appropriate scale, 
use images and metaphors to gain/retain attention but dont arouse undue emotion, 
assume low numeracy of a general public audience and adopt a less-is-more approach by reducing the need for inference and making clear and explicit comparison, 
be cautious about interactivity and animations (may introduce unnecessary complexity), 
avoid chart junk (like 3d bar charts),

@@

## Types of Uncertainty Visualisation Papers
Uncertainty visualisation papers can be organised according to the *goal* of the experiment. Evaluation experiments are the standard rule for visualisations because the human brain is not as reliable as mathematical calculation. Therefore, user studies often aim to assess the limitations, biases, and heuristics of our mental calculator so that we can better understand the problems we may encounter when we plot our data. This is not to say any paper that suggests a visualisation without an evaluation experiment is completely lacking in justification, and there are many papers that suggest a novel visualisation without an evaluation study. Sometimes these papers are a preliminary step in finding a solution for common problems and intend to evaluate the visualsiation in later work. Until that later work is done, it is often difficult to accept one particular representation. While justification could come in the form of discussing established concepts in visualisation, such as the hierarchy of perceptual tasks, even these may need an evaluation study to back up their claims.

@@

**Aside: I should maybe change this to relatability to uncertainty, value extraction, decision making and trust, and meta papers or something. Maybe perception and evaluation papers?**

@@

## Uncertainty and Inference
When we consider mapping uncertainty, a reasonable questions that should come to mind is "uncertainty about what?". Every single thing concept or activity in statistics does not have uncertainty. For example, descriptive statistics cannot have uncertainty as they are just describing statistics as they are, the sample mean is the sample mean, we are not uncertain about its value. A sample variance describes dispersion, which is also how we describe uncertainty, but the sample variance in of itself does not have any uncertainty. Most of the mathematical uncertainty concepts have this idea embedded in them. For example, a confidence interval is an interval drawn around the sample mean that will theoretically contain the *populaton mean* 95% of the time. This issue has been noticed several times, however it is often brought up as a *task* dependence rather than an understanding that uncertainty is inherrently linked to inference on a particular statistic. @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. @Munzner2009 created a nested model for visualisation that highlighted how the first mistake that can be made in a visualisation is in the problem characterisation, and failling to do it well can cause downstream effects and damage the effectivness of a visualisation. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it. Health have a wide range of metrics used to communicate different risks [@@Spiegelhalter2017] indicating they utilise new statistics based off the specific aspect of risk that is relevant to the issue. This should make it clear that uncertainty only exists when we are performing inference. Therefore it only makes sense to describe uncertainty in terms of something to be uncertain *about*, it is not some a latent feature of statistics and data.

@@

**Aside: Include table that shows broad uncertainty definition and then PDF visualisation**

@@

## Evaluation Experiments and The Grammar of Graphics
The first issue in uncertainty visualisation is there does not seem to be a clear rule for what makes a particular visualisation a depiction of uncertainty. This problem seems to start at the lack of the mathematical definition of uncertainty which then bleeds into a lack of a definition for an "uncertainty visualisation". For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in a large number of uncertainty visualisation experiments. How can a chart omit uncertainty according to the grammar of graphics, but be used to describe it in several evaluation studies? The answer to this question is related to the definition problem discussed earlier, but to understand how, we will need to go through the grammar of graphics, and its absence in the uncertainty visualisation literature.

**Aside: I think I want the "in order to visualise uncertainty you need to quantify it and the quantification does not align with the definition" point here since this section is now about value extraction.**

@@

@@

Computational equivalency in visualisation is also something that has been well researched. For simple tasks such as value extraction, there is a hierarchy to perceptual tasks where extracting visual information in some forms is easier than others. The hierarchy established 40 years ago by @Cleveland1984 is: 
1) position along a common scale
2) position along a non-aligned scale
3) length
4) angle/slope
5) area
6) volume
7) colour
This hierarchy has had additions and adjustments over the years. For example, this hierarchy is a general rule, but the hierarchy can change from individual to individual. Additionally, there are other graphical rules to consider such as guestalt heuristics, broader methods of extraction, and attention principles. These established visualisation concepts allow us to anticipate the ease with which certain pieces of information will be extracted from a plot.

@@

**Aside: INCLUDE A GRAPHIC THAT SHOWS WHERE EACH TYPE OF UNCERTAINTY VISUALISATION SITS IN THE GRAMMAR OF GRAPHICS AND HOW DIFFERENT THEY ARE ACROSS MULTIPLE STUDIES**

@@

There is also the common view that uncertainty as a dimensionality problem. This perspective is not quite right. It is important to keep in mind that just because information is in a graphic, that does not mean it will be "seen". The phenomena of inattentional blindness shows that there is no perception without attention and it is powerful enough that participants can fail to be aware of random objects appearing on a screen or a gorilla walking through a basketball game [@simons1999gorillas; @mack2003inattentional]. Therefore if information is mapped to graphical elements that are so low on hierarchy they can be ignored, they might as well not be there at all. The people who don't plot uncertainty because they think its unimportant and the people who relegate uncertainty to the lowest ranking aesthetic on the list are both saying the same thing, "I think the uncertainty is unimportant". Including uncertainty is worth very little if no attention is left to see it. This does not mean we cannot put a *large* amount of information in a graphic. Glyph maps can be used to depict the multi-dimensional information in spatial-temporal data by mapping line plots to map locations, but we still need to decide what information is important. The map can either present trends in the global or local variance depending on whether or not the line plot is scaled globally or locally [@Wickham2012], however smaller details are almost impossible to convey. No matter how much information we try and put in a plot, there will always be only a handful of key takeaways.

@@

## Noise and Signal in Uncertainty Visualisation
Uncertainty is noise. This may be an obvious statement, but what is not obvious is that the existing literature sees noise as a signal. When we directly ask "What is the varinance of this variable" the variance *becomes* the signal. Noise and signal are not an inherrent property of variance, or meta-data or the aspects of an analysis that are usually secondary considerations, it is a property of the data that *promotes* a message we are trying to infer and the data that *supresses* it. The replication problem in uncertainty visualisation, where experiments are simply repeating established visualisation work, comes from the misunderstandings established in previous sections. Uncertainty is poorly defined, therefore authors do not understand it is not an inherrent property of the data and is actually related to a particular inference question, this incorrect view that uncertainty is always variance or probabilities leads to the class of "uncertainty visualisations" that depict different mathematical objects or visual features with little in the sense of consistent rules, this complicated class has caused visualisation authors to miss that they are simply repeating the established evaluation results from normal information visualisations. This situation is a bit of a mess. It also highlights a unique and facinating problem faced by uncertainty visualisation. If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?

@@

# Great Examples
Of course we do not believe every uncertainty visualisation is unusable and every experiment is build upon misguided assumptions. Despite the common problems detialled in the previous sections, there is some interesting work in the uncertainty visualisation space. This can come in the form of an uncertainty visualisation that identifies or fixes pitfalls in the previous sections (without identifying them) or does an experiment that identifies useful information that avoids these pitfalls.

@@

One example of a great uncertainty visualisation method that was designed with signal supression in mind is the value supressing uncertainty pallet (VSUP). Depicts a bivariate map which uses a bivariate colour palette that is created by blending two single hue colour palettes. One colour represents the variable of interest while the other represents the sampling error of that variable. There are two immediate problems with this method. First of all, uncertainty is being expressed with hue and saturation which @Maceachren2012 found to be the worst aesthetics to map to uncertainty to as they don't have an intuitive interpretation. Value has a natural connection to uncertainty (lighter values equate to higher uncertainty and darker values equate to more certainty) so it is a much more appropriate choice. While the `Vizumap` data does depict areas of light and darkness, they are largely irrelevant to the uncertainty measure causing our heuristics to lead us to the incorrect conclusions. The Value-Suppressing Uncertainty Palettes (VSUP) shown in @fig-vsup maps estimates to the hue and error to the value thereby creating a more intuitive plot [@Correll2018]. Additionally, at high levels of uncertainty VSUP only has one output colour, which prevents viewers from decrypting any particular value and also avoids enforcing a binary encoding of significance [@Correll2018]. Unfortunately VSUP are not easy to combine with packages like `Vizumap` which leaves it still somewhat difficult to express this encoding in practice, however the combination of a bivariate map with VSUP has shown to improve decisions in the face of uncertainty [@Correll2018].

>>>>

---
title: "Noisy Work: A Review of The Uncertainty Visualisation Literature"
author: Harriet Mason
bibliography: references.bib
date: last-modified
toc: true
number-sections: true
format: pdf
editor_options: 
  chunk_output_type: console
abstract: In reviewing the literature of uncertainty visualisation
---

# Introduction

> Edit note: this section is untouched since the confirmation. Might need a rewrite

@@

@@

# Similar work
> Edit note: I think with the addition of more summary papers I should just summarise common themes of issues the papers ran into instead of detailling the work paper by paper.

There are a wealth of uncertainty communication and visualisation literature reviews that already exist. Interestingly they all seem to have identified similar pitfalls or fallen into one themselves.

@@

## Coloquial Definitions
The most common definition of uncertainty used in papers that discuss visualising it are colloquial. Most definitions make some nod towards variance, precision, trust, error, missing values or another related concepts. Since *uncertainty*, unlike *probability*, does not have a strict definition that is established in mathematics, there is large variety in its usage. Some consider it to be synonymous with probabiltiy, some consider it to be an encompassing umbrella term of which probability is only an element, and some consider it to by something else entirely. This lack of a definition leads to uncertainty being swapped out freely with other vague words such as ambiguity and risk. Definitions presented in papers that discuss uncertainty visualisation want it to be encompassing and include everything a layperson might consider when trying to consider "uncertainty". The issue with these broad definitions is is that in order to visualize something, it needs to be quantified, and in order to quantify something, we need to know exactly what it is we are counting. This discrepancy appears regularly in the uncertainty visualisation literature. Often a vague definition such as "uncertainty is anything that isn't deterministic", or even a refusal to define uncertainty at all in an experiments introduction will be accompanied by a highly specific quantification of uncertainty as a probability density distribution of a forecast when providing the visualisation. Ironically it seems that many of the papers that refer to themselves as "uncertainty" papers are just suffering from a form of uncertainty, a lack of precision.[^1]

@@

## Taxonomy Definitions
### Taxonomies of Uncertainty
> Edit Note: A lot of this is going to be moved to the uncertainty definition specific section of the vis lit review and the point here should just be "this is a mess" instead of trying to find any reason in it. My opinion and goal for this section has changed a lot since writing this (specifically deciding that the encapsulating taxonomy doesnt fix the problem and a maths definition is needed) so it is one of the sections that needs big rewrites.

@@

> Edit note: I dont think I will keep this figure in the paper. The goal was to show they are inconsistent but I think a sentance or two in the text will achieve that.

@@

![*I need to fix the taxonomy illustration because inputs and model uncertainty need to be swapped*. Depicts an illustration of the taxonomy described in @utypo. From right to left the drawing shows the location, level and nature of uncertainty with examples of that category underneath. A specific source of uncertainty from the location can be mapped to a level of ignorance that can increase or decrease (i.e. moving up or down the green line) depending on the nature of the uncertainty. Identifying the location, level and nature of your uncertainty allows you to better understand it. ](taxonomyvis.jpeg){#fig-taxonomy}

> Edit note: Everything from this point down was used to try and provide some justification for why this dimensionality concept could work, but I have since changed my mind and am probably going to get rid of this section unless it fits somewhere else.

@@

Mapping uncertainty into the 3D space defined by location, level, and nature can expand beyond the narrow case of statistical modelling it was defined for. *(find the citation for this)* included "consensus" as a taxonomy element, however it was not included in the final definition, but it can easily be added as a location that would align with the communication stage of the data pipeline. Several authors [@Spiegelhalter2017] *(I have other papers)* considered the method by which uncertainty was quantified (e.g. by a single value, a function, simulated outcomes, etc), which can be be thought of as a more precise categorisation of the level axis. This taxonomy also ignores contextual information about the data, such as spatial or temporal considerations. This flexibility means that the use of this framework is not limited by a specific case and should be used going forward to define what aspect of the uncertainty space a contribution exists in.

@@

### Taxonomies of Visual Uncertainty
> Edit Notes: Similar to the "similar works" section, I think this section would be better off with a short sentance about each paper and comments on their similarities because a paragraph for each paper is too much

There are some taxonomies of visual uncertainty. These taxonomies seem to seek to define "uncertainty visualisation" and a combined term rather than as two separate words. We largely disagree with this conceptualization of uncertainty visualisations as it blurs the line between the mathematical estimations and assumptions (the uncertainty part) and the visual depiction of those mathematical objects (the visualisation part). While we will not focus on these taxonomies, it is still important to mention them and highlight the reasons they are not good practice.

@@

## Contextual Information
> Edit Note: This section should probably be moved to the visualisation specific section, although that depends on the tone that I end up with after the rewrites of the earlier sections.

Some uncertainty work, specifically uncertainty visualisation, focus on the context surrounding the uncertainty. These are things like spatial data, temporal data, category data, etc. While other discussions of uncertainty may consider this distinction more relevant to their work, here we will consider this contextual information. Contextual information is important for interpretation and understanding of graphics, but it does not in of itself generate uncertainty. This is not unique to the properties of spatial information, other statistics information that is important for modelling and assumptions, such as non-linearity, class levels, etc do not contribute to uncertainty. Operations on data, inherrent randomness, and assumptions that create bias generate uncertainty, which is what we are talking about here. A confidence interval is a confidence interval whether it came from spatial data, temporal, or cross sectional data.

@@

# General Attitude Towards Uncertainty (Uncertainty Communication Boradly)

> Edit note: this section needs a huge rewrite. The idea of this section blends way too much with the measurement discussions in the "noise and signal" section. I have already taken paragraphs from here and just pasted it in the noise section so it might not be coherrent. Additionaly this section goes on SEVERAL long rambles that realistically should have been a sentance or not included at all.

It would be impossible to discuss uncertainty visualisation without mentioning the broader conversation around uncertainty communication. The conversation around uncertainty communication is often centered around the psychology is communicating risks.

@@

@@

## Types of Uncertainty Visualisation Papers

> Edit notes: Originally I was planning to have the discussion following this section relate to these "types" of papers, but that didn't end up happening. Depending on what I do with later sections, this might need a rewrite or to just be removed, since I'm not sure how relevant this is to the main ideas of the paper.

Uncertainty visualisation papers can be organised according to the *goal* of the experiment. Evaluation experiments are the standard rule for visualisations because the human brain is not as reliable as mathematical calculation. Therefore, user studies often aim to assess the limitations, biases, and heuristics of our mental calculator so that we can better understand the problems we may encounter when we plot our data. This is not to say any paper that suggests a visualisation without an evaluation experiment is completely lacking in justification, and there are many papers that suggest a novel visualisation without an evaluation study. Sometimes these papers are a preliminary step in finding a solution for common problems and intend to evaluate the visualsiation in later work. Until that later work is done, it is often difficult to accept one particular representation. While justification could come in the form of discussing established concepts in visualisation, such as the hierarchy of perceptual tasks, even these may need an evaluation study to back up their claims.

@@

@@

## Defining uncertainty with resepct to inference
When we consider mapping uncertainty, a reasonable questions that should come to mind is "uncertainty about what?". Every single thing concept or activity in statistics does not have uncertainty. For example, descriptive statistics cannot have uncertainty as they are just describing statistics as they are, the sample mean is the sample mean, we are not uncertain about its value. A sample variance describes dispersion, which is also how we describe uncertainty, but the sample variance in of itself does not have any uncertainty. Most of the mathematical uncertainty concepts have this idea embedded in them. For example, a confidence interval is an interval drawn around the sample mean that will theoretically contain the *populaton mean* 95% of the time. This issue has been noticed several times, however it is often brought up as a *task* dependence rather than an understanding that uncertainty is inherrently linked to inference on a particular statistic. @Wallsten1997 argue that the best method for evaluating or combining subjective probabilities depends on the uncertainty the decision maker wants to represent and why it matters. @Munzner2009 created a nested model for visualisation that highlighted how the first mistake that can be made in a visualisation is in the problem characterisation, and failling to do it well can cause downstream effects and damage the effectivness of a visualisation. @Fischhoff2014 looks at uncertainty visualisation for decision making decides that we should have different ways of communicating uncertainty based off what the user is supposed to do with it. Health have a wide range of metrics used to communicate different risks [@@Spiegelhalter2017] indicating they utilise new statistics based off the specific aspect of risk that is relevant to the issue. This should make it clear that uncertainty only exists when we are performing inference. Therefore it only makes sense to describe uncertainty in terms of something to be uncertain *about*, it is not some a latent feature of statistics and data.

@@

> Edit notes: Need some way to show uncertainty visualisation papers provide sweeping definitions and then visualise a simple PDF everytime.

@@

## Uncertainty evaluation experiments repeat visualisation studies
The first issue in uncertainty visualisation is there does not seem to be a clear rule for what makes a particular visualisation a depiction of uncertainty. This problem seems to start at the lack of the mathematical definition of uncertainty which then bleeds into a lack of a definition for an "uncertainty visualisation". For example @Leland2005 mentions that popular graphics, such as pie charts and bar charts omit uncertainty, however at least one or both of these charts are used in a large number of uncertainty visualisation experiments. How can a chart omit uncertainty according to the grammar of graphics, but be used to describe it in several evaluation studies? The answer to this question is related to the definition problem discussed earlier, but to understand how, we will need to go through the grammar of graphics, and its absence in the uncertainty visualisation literature.

> Edit notes: The "in order to visualise uncertainty you need to quantify it and the quantification does not align with the definition" point could also work here since this section is now about value extraction.

@@

> Edit notes: Grammar of graphics is just a placeholder, Di suggested to replace this with my own pipeline visualisation based on some data wrangling papers.

@@

Computational equivalency in visualisation is also something that has been well researched. For simple tasks such as value extraction, there is a hierarchy to perceptual tasks where extracting visual information in some forms is easier than others. The hierarchy established 40 years ago by @Cleveland1984 is:  
1) position along a common scale. 
2) position along a non-aligned scale. 
3) length. 
4) angle/slope. 
5) area. 
6) volume. 
7) colour

> Edit notes: need to add the notes from the susan lit review that has the more recent perceptual task list

This hierarchy has had additions and adjustments over the years. For example, this hierarchy is a general rule, but the hierarchy can change from individual to individual. Additionally, there are other graphical rules to consider such as guestalt heuristics, broader methods of extraction, and attention principles. These established visualisation concepts allow us to anticipate the ease with which certain pieces of information will be extracted from a plot.

@@

> Edit note: need to include a graphic that illustrates the wide spread of mathematical information and visual techniques that are visualised in these experiments.

@@

There is also the common view that uncertainty as a dimensionality problem. This perspective is not quite right and the literature seems to be about 50/50 on whether uncertainty is a psychological or dimensionality problem. @Kinkeldey2014 also identified this issue when they highlighted that the literature makes it unclear if uncertainty is a variable in of itself, something that should be interpreted with the main variable, or if it is metadata. It is important to keep in mind that just because information is in a graphic, that does not mean it will be "seen". The phenomena of inattentional blindness shows that there is no perception without attention and it is powerful enough that participants can fail to be aware of random objects appearing on a screen or a gorilla walking through a basketball game [@simons1999gorillas; @mack2003inattentional]. Therefore if information is mapped to graphical elements that are so low on hierarchy they can be ignored, they might as well not be there at all. The people who don't plot uncertainty because they think its unimportant and the people who relegate uncertainty to the lowest ranking aesthetic on the list are both saying the same thing, "I think the uncertainty is unimportant". Including uncertainty is worth very little if no attention is left to see it. This does not mean we cannot put a *large* amount of information in a graphic. Glyph maps can be used to depict the multi-dimensional information in spatial-temporal data by mapping line plots to map locations, but we still need to decide what information is important. The map can either present trends in the global or local variance depending on whether or not the line plot is scaled globally or locally [@Wickham2012], however smaller details are almost impossible to convey. No matter how much information we try and put in a plot, there will always be only a handful of key takeaways.

@@

## Measuring uncertainty as noise, not signal
Uncertainty is noise. This may be an obvious statement, but what is not obvious is that the existing literature sees noise as a signal. When we directly ask "What is the varinance of this variable" the variance *becomes* the signal. Noise and signal are not an inherrent property of variance, or meta-data or the aspects of an analysis that are usually secondary considerations, it is a property of the data that *promotes* a message we are trying to infer and the data that *supresses* it. The replication problem in uncertainty visualisation, where experiments are simply repeating established visualisation work, comes from the misunderstandings established in previous sections. Uncertainty is poorly defined, therefore authors do not understand it is not an inherrent property of the data and is actually related to a particular inference question, this incorrect view that uncertainty is always variance or probabilities leads to the class of "uncertainty visualisations" that depict different mathematical objects or visual features with little in the sense of consistent rules, this complicated class has caused visualisation authors to miss that they are simply repeating the established evaluation results from normal information visualisations. This situation is a bit of a mess. It also highlights a unique and facinating problem faced by uncertainty visualisation. If asking direct questions about uncertainty causes us to treat it as a signal, how do we evaluate uncertainty as *noise*?

@@

# Great Examples

> Edit notes: this section seems to go off a bit on dimensionality of visual tasks and stuff that might be better put in the noise and signal section about measuring uncertainty.

Of course we do not believe every uncertainty visualisation is unusable and every experiment is build upon misguided assumptions. Despite the common problems detialled in the previous sections, there is some interesting work in the uncertainty visualisation space. This can come in the form of an uncertainty visualisation that identifies or fixes pitfalls in the previous sections (without identifying them) or does an experiment that identifies useful information that avoids these pitfalls.

Uncertainty that is created by steps that are further down the pipeline are often ignored by the uncertainty visualisation literature, however there are visualisations that make these considerations. If uncertainty had a more encapsulating mathematical definition, the uncertainty created at earlier stages might be able to be visualised with the model uncertainty that is typically expressed through confidence intervals. That being said, even with a definition that allows us to combine these uncertainties into a single value, it still might be desirable to see which stage of the analysis is creating uncertainty in the visualisaiton. One example of a visualisation that includes that information is visualisations of climate scenario uncertainty, as shown in @fig-climatescenario. Uncertainty is introduced into this model by the choices of individuals which is difficult to model or predict and is also unlikely to be constant over time. Therefore specific scenarios are modelled and shown, each with their own model uncertainty. This highlights a method that can be used to visualise uncertainty from earlier in the pipeline that is often forgotten or ignored via assumptions.

![I will just make my own version of a climate scenario plot, this one is just here from a random site as a place holder](climatescenario.png){fig-climatescenario}

@@

There are a number of visualisation methods that do take the approach of uncertainty as signal suppression. These visualisation examples try to establish uncertainty as something that should be interpreted *with* the signal and take several approaches to do so. @fig-maps depicts the array of approaches that have been used to visualise uncertainty in a map.

The first approach to embedding noise in the signal is to adjust the signal visualisation so that it includes the noise. This method has been particularly popular with map visualisations. To visualise uncertainty such that it does its job of signal supression the uncertainty and the signal need to be interpreted in conjunction with each other. There are several suggested ways to go about this, most of which assume the "no uncertianty" version of the visualisation is a simple choropleth map. There are two methods to include uncertainty in a map. The first is to include uncertainty as error and map our signal and noise to a two dimensional *colour* space, which is the method adopted by the bivariate map and the value supressing uncertainty pallet [@Correll2018]. These methods rely on the assumption that colour hue and value exist mentally as a two dimensional colour space and that distance in this colour space is linear (or at least appears linear using our scale). However, for uncertainty to be interpreted in the visualisation *with* the signal, our brain would need to be able to flatted this two dimensional space into a single space of "signal validity", and there is simply not evidence that we can do this. This visualisation method brings up an interesting concept though. If there are two aesthetics we can use that are flattened into a single channel in our mind, then plotting measurement and error to each of those aesthetics is a viable method to visualise uncertainty. The idea of two perceptual tasks flattening into one variable in the mind of the viewer may be wishful thinking, but it is not impossible given we are not certain on how the perceptual tasks are mapped within the human brain. @Sterzik2023 found that when a value was mapped to the textures of stippling, hatching, and triangles, and found that the difference between two points on this one dimensional texture was actually a 2D space (likely "business" and light/darkness). This idea of two aesthetics flattening into one cohesive variable in the perceptual space is particularly interesting for uncertainty visualisation.

An alternative method to visualising uncertainty in a map is to combine uncertainty and signal at the earlier stage so the "supressed signal" is represented by a single variable. This statistic can then be expressed in a one dimensional colour space, which is a method addopted by the Baysian surprise metric map [@Ndlovu2023] and the excedance probability map [Lucchesi2017]. These methods do not require us to compress colour hue and value into a single variable to interpret signal and noise as a single variable, it is done explicitly for us before the visualisaiton. These methods require more specificity in *how* the uncertainty contains important information.

::: {#fig-maps}

![](mappals/img1.png)

![](mappals/img2.png)

![](mappals/img4.png)
![](mappals/img3.png)

![](mappals/img5.png)

*I am going to replace this with my own R coded maps so they are all on the same data and signal/noise.* Four different ways to include error information in a map. A typical chloropleth map is considered the map that ignores uncertainty and simply visualises the signal. The bivariate map visualises the uncertainty, but does so using two different visual channels, which gives the perception of two different signals (or one signal in the assumptions the colors contrast with each other). An adjustment to the bivariate map is the VSUP which is similar to the bivariate plot except it merges high uncertainty values in the palette to reduce our ability to discern values that have a high enough error to not be a valid signal. The Bayesian surprise metric is similar to the lineup protocol, where the colour value is the difference between the observers posterior and prior beliefs. The exceedance probability map is a direct visualisation of the probability of a specific hypothesis.
:::

These maps make the importance of combining uncertainty and signal in a single visual channel clear. A chloropleth map will show signal that is not valid inference because of high uncertainty. At the other end of the spectrum, the bivariate map will show signal that is not always interesting because it forces us to interpret uncertainty and signal differently. The value supressing uncertainty pallet was clearly designed with hypothesis testing in mind, because it combines highly uncertain values. The main difference between the VSUP and the bivariate colour map is actually the subtle combination of signal and noise through discernability. In this sense, the uncertainty visualisation maps that combine uncertainty and signal before it is conveyed visually simplify the visualisation and make it easier to interpret value supressed signal. This is what is done by the exceedence probability map and the Bayesian surprise metric map. They are more explicit in how the uncertainty should supress the signal since they make the specific hypothesis explicit.

Another method that has been used is to visualise samples from the distribution we are trying to draw inference on instead of estimates along with a variance or error. This has been used in maps with the pixel map [Lucchesi2017], but is more commonly used in animation with the HOPs plots [@Hullman2015] or similar concepts [@Blenkinsop2000] . The difficulty with resampling methods is that they are typically conflated with animation even though this does not necessarily need to be true, as can be seen in the @fig-pixelmap. Even though @Kale2018 found that users were more sensitive to the underlying trend when shown the HOPs plot over the static outcoms plot, this effect went away when the speed of the animation increased, indicating that performance difference might be due to overplotting in the static case and a smaller sample size in the static condition would produce the same result. Therefore, so long as the sample size of the number of outcomes is kept to a level where individual samples are still visible, there is no evidence that an outcome plot *needs* to be animated. Visualisations that opt to express a signal as a sample rather than an estimate have the potential to supress signal since it is not explicitly visualised, however this has yet to be shown in evaluation experiments. This is not to say that visualisations of mass would not be able to perform signal supression, but a sample can easily be expressed using aesthetics such as colour on a map and mass visualisation often struggling with issues such as over or under smoothing.

![](pixelmap.png){fig-pixelmap}
